---
title: "Signal Detection Theory"
author: "Julia Haaf & Stephen Rhodes"
output:
  ioslides_presentation:
    logo: pictures/MUlogoRGB.png
    widescreen: true
subtitle: The Real Deal
---

## Signal Detection Experiment | Example

>- Subliminal perception
>- Very brief presentation of a word vs. no word
>- Signal: + ... TRUMP ... *****
>- Noise: + ... &emsp; &emsp; &emsp; ... *****

## Signal Detection Experiment | Example

**Results**

| Stimulus | Present response | Absent Response | Total | 
|:------|:-----:|:---------:|:------:| 
|   Signal  |  75  |    25   |    100  | 
|  Noise  |  30 |   20   |   50  | 
|  Total  |  105 |   45   |     | 

>- Hits, Misses, False Alarms, Correct Rejections
>- vs. Accuracy

## Signal Detection Theory vs. Signal Detection Experiment

>- Signal Detection Theory (SDT): Model-based method of assessing performance
>- Originally: Developed for signal detection experiments
>- Now: Used for perception, memory, decision theory, attitude research, ...
>- Originally: Two-choice task
>- Now: Any form of forced-choice task including rating scales, continuous measures, ...

## Signal Detection Theory

Models are a bit more complicated and rely on some knowledge of continuous random variables.

```{r, echo = F}
x <- seq(-3, 3, .01)
y <- dnorm(x)
plot(x, y, yaxt = 'n', type = "l", lwd = 2, ylab = "")
```


# Probability Theory 101

## Density function

The density function (pdf) of a continuous random variable $X$ is a function such that the area under the function between $a$ and $b$ corresponds to the probability $Pr(a < x \leq b)$. The value of $f(x)$ is the probability density.

## Density function

The density function (pdf) of a continuous random variable $X$ is a function such that the area under the function between $a$ and $b$ corresponds to the probability $Pr(a < x \leq b)$. The value of $f(x)$ is the probability density.

**Example**

```{r, echo = F, fig.height=2.5, fig.asp = .5, fig.align='center'}
par(mar = c(4, 4, 0, 2))
plot(x, y, type = "l", lwd = 2
     , ylab = "Density f(X)", xlab = "X")
text(2.3, .38, "pdf of a normal", cex = 1.3)
```

## Density function

The density function (pdf) of a continuous random variable $X$ is a function such that the area under the function between $a$ and $b$ corresponds to the probability $Pr(a < x \leq b)$. The value of $f(x)$ is the probability density.

**Example**

```{r, echo = F, fig.height=2.5, fig.asp = .5, fig.align='center'}
par(mar = c(4, 4, 0, 2))
plot(x, y, type = "l", lwd = 2
     , ylab = "Density f(X)", xlab = "X")
text(2.3, .38, "pdf of a normal", cex = 1.3)

make.line <- function(x, mu = 0, s2 = 1, ...){
  y <- dnorm(x, mu, sqrt(s2))
  lines(x = rep(x, 2), y = c(0, y) , ...)
}

make.poly <- function(a, b, mu = 0, s2 = 1, seq.by = .01, ...){
  y <- dnorm(seq(a, b, seq.by), mu, sqrt(s2))
  xx <- c(seq(a, b, seq.by), seq(b, a, -seq.by))
  yy <- c(y, rep(0, length(y)))
  polygon(xx, yy, border = NA, ...)
}

make.poly(-1.2, .3, col = "cornflowerblue")
make.line(-1.2, lwd = 2, col = 1)
make.line(.3, lwd = 2)
lines(x, y, lwd = 2)
text(-1.2, .4, "a", cex = 1.2)
text(.3, .4, "b", cex = 1.2)
```

## Density function in R

```{r}
x <- seq(-3, 3, .01)
head(x)
y <- dnorm(x = x, mean = 0, sd = 1)
head(y)
```

## Density function in R

```{r, fig.height=4, fig.width = 6, fig.align='center'}
x <- seq(-3, 3, .01)
y <- dnorm(x = x, mean = 0, sd = 1)

plot(x, y, type = "l")
```

## Cumulative Distribution Function

Let $F$ denote the cumulative distribution function (CDF) of random variable $X$. Then,
\[F(x) = Pr(X \leq x)\].

## Cumulative Distribution Function

Let $F$ denote the cumulative distribution function (CDF) of random variable $X$.
\[F(x) = Pr(X \leq x)\].

**Example**

```{r, echo = F, fig.height=3, fig.width = 8,  fig.align='center'}
par(mar = c(4, 4, .2, 2))
layout(matrix(1:2, ncol = 2))
plot(x, y, type = "l", lwd = 2
     , ylab = "Density f(X)", xlab = "X")
text(2.5, .38, "pdf", cex = 1.3)

make.line <- function(x, mu = 0, s2 = 1, fun = "pdf", ...){
  if(fun == "pdf"){
    y <- dnorm(x, mu, sqrt(s2))
  }else{
    y <- pnorm(x, mu, sqrt(s2))
    }
  lines(x = rep(x, 2), y = c(0, y) , ...)
}

make.poly <- function(a, b, mu = 0, s2 = 1, seq.by = .01, fun = "pdf", ...){
  if(fun == "pdf"){
    y <- dnorm(seq(a, b, seq.by), mu, sqrt(s2))
  }else{
    y <- pnorm(seq(a, b, seq.by), mu, sqrt(s2))
    }
  xx <- c(seq(a, b, seq.by), seq(b, a, -seq.by))
  yy <- c(y, rep(0, length(y)))
  polygon(xx, yy, border = NA, ...)
}

make.poly(-3, .3, col = "cornflowerblue")
make.line(.3, lwd = 2)
lines(x, y, lwd = 2)
text(.3, .4, "x", cex = 1.2)

plot(x, pnorm(x), type = "l", lwd = 2
     , ylab = "Cummulative Density", xlab = "X")

make.poly(-3, .3, col = "cornflowerblue", fun = "cdf")
make.line(.3, lwd = 2, fun = "cdf")
lines(x, pnorm(x), lwd = 2)
text(.3, .8, "x", cex = 1.2)
text(-.5, .1, "F(x)", cex = 1.2)
```

## Using CDF and pdf

```{r, echo = F, fig.height=4, fig.width = 7, fig.align='center'}
par(mar = c(4, 4, 0, 2))
plot(x, y, type = "l", lwd = 2
     , ylab = "Density f(X)", xlab = "X")
text(2.3, .38, "pdf of a normal", cex = 1.3)

make.line <- function(x, mu = 0, s2 = 1, ...){
  y <- dnorm(x, mu, sqrt(s2))
  lines(x = rep(x, 2), y = c(0, y) , ...)
}

make.poly <- function(a, b, mu = 0, s2 = 1, seq.by = .01, ...){
  y <- dnorm(seq(a, b, seq.by), mu, sqrt(s2))
  xx <- c(seq(a, b, seq.by), seq(b, a, -seq.by))
  yy <- c(y, rep(0, length(y)))
  polygon(xx, yy, border = NA, ...)
}

make.poly(-1.2, .3, col = "cornflowerblue")
make.line(-1.2, lwd = 2, col = 1)
make.line(.3, lwd = 2)
lines(x, y, lwd = 2)
text(-1.2, .4, "a", cex = 1.2)
text(.3, .4, "b", cex = 1.2)
```

**What is $Pr(a < x \leq b)$**

## Using CDF and pdf

```{r, echo = F, fig.height=4, fig.width = 7, fig.align='center'}
par(mar = c(4, 4, 0, 2))
plot(x, y, type = "l", lwd = 2
     , ylab = "Density f(X)", xlab = "X")
text(2.3, .38, "pdf of a normal", cex = 1.3)

make.poly(-3, .3, col = "cornflowerblue")
# make.poly(-3, -1.2, col = adjustcolor("mediumorchid", .5))
# make.line(-1.2, lwd = 2, col = 1)
make.line(.3, lwd = 2)
lines(x, y, lwd = 2)
# text(-1.2, .4, "a", cex = 1.2)
text(.3, .4, "b", cex = 1.2)
```

## Using CDF and pdf

```{r, echo = F, fig.height=4, fig.width = 7, fig.align='center'}
par(mar = c(4, 4, 0, 2))
plot(x, y, type = "l", lwd = 2
     , ylab = "Density f(X)", xlab = "X")
text(2.3, .38, "pdf of a normal", cex = 1.3)

make.line <- function(x, mu = 0, s2 = 1, ...){
  y <- dnorm(x, mu, sqrt(s2))
  lines(x = rep(x, 2), y = c(0, y) , ...)
}

make.poly <- function(a, b, mu = 0, s2 = 1, seq.by = .01, ...){
  y <- dnorm(seq(a, b, seq.by), mu, sqrt(s2))
  xx <- c(seq(a, b, seq.by), seq(b, a, -seq.by))
  yy <- c(y, rep(0, length(y)))
  polygon(xx, yy, border = NA, ...)
}

make.poly(-1.2, .3, col = "cornflowerblue")
make.line(-1.2, lwd = 2, col = 1)
make.line(.3, lwd = 2)
lines(x, y, lwd = 2)
text(-1.2, .4, "a", cex = 1.2)
text(.3, .4, "b", cex = 1.2)
```

**What is $Pr(a < x \leq b)$**

## Using CDF and pdf

```{r, echo = F, fig.height=3, fig.width = 4.5, fig.align='center'}
par(mar = c(4, 4, 0, 2))
plot(x, y, type = "l", lwd = 2
     , ylab = "Density f(X)", xlab = "X")
# text(2.3, .38, "pdf of a normal", cex = 1.3)

make.poly(-3, .3, col = "cornflowerblue")
make.poly(-3, -1.2, col = adjustcolor("mediumorchid", .5))
make.line(-1.2, lwd = 2, col = 1)
make.line(.3, lwd = 2)
lines(x, y, lwd = 2)
text(-1.2, .4, "a", cex = 1.2)
text(.4, .4, "b", cex = 1.2)
```

\[Pr(a < x \leq b) = F(b) - F(a)\]


## Using CDF and pdf

```{r, echo = F, fig.height=3, fig.width = 4.5, fig.align='center'}
par(mar = c(4, 4, 0, 2))
plot(x, y, type = "l", lwd = 2
     , ylab = "Density f(X)", xlab = "X")
# text(2.3, .38, "pdf of a normal", cex = 1.3)

make.poly(-3, .3, col = "cornflowerblue")
make.poly(-3, -1.2, col = adjustcolor("mediumorchid", .5))
make.line(-1.2, lwd = 2, col = 1)
make.line(.3, lwd = 2)
lines(x, y, lwd = 2)
text(-1.2, .4, "a", cex = 1.2)
text(.4, .4, "b", cex = 1.2)
```

\[Pr(a < x \leq b) = F(b) - F(a)\]


```{r}
pnorm(.3, 0, 1) - pnorm(-1.2, 0, 1)
```

## Quantile Function

The $p^{th}$ quantile of a distribution is the value $q_p$ such that $Pr(X \leq q_p) = p$.

>- Reverse of the CDF

## Quantile Function

The $p^{th}$ quantile of a distribution is the value $q_p$ such that $Pr(X \leq q_p) = p$.

**Reverse of the CDF**

```{r, echo = F, fig.height=3, fig.width = 4.5, fig.align='center'}
par(mar = c(4, 4, .2, 2))
plot(x, y, type = "l", lwd = 2
     , ylab = "Density f(X)", xlab = "X")
# text(2.3, .38, "pdf of a normal", cex = 1.3)

make.poly(-3, qnorm(.75), col = adjustcolor("mediumorchid", .5))
make.line(qnorm(.75), lwd = 2, col = 1)
lines(x, y, lwd = 2)
text(-.5, .15, "p = .75", cex = 1.2)
text(qnorm(.75), .38, "?", cex = 1.2)
```

## Quantile Function

The $p^{th}$ quantile of a distribution is the value $q_p$ such that $Pr(X \leq q_p) = p$.

```{r, echo = F, fig.height=3, fig.width = 4.5, fig.align='center'}
par(mar = c(4, 4, .2, 2))
plot(x, y, type = "l", lwd = 2
     , ylab = "Density f(X)", xlab = "X")
# text(2.3, .38, "pdf of a normal", cex = 1.3)

make.poly(-3, qnorm(.75), col = adjustcolor("mediumorchid", .5))
make.line(qnorm(.75), lwd = 2, col = 1)
lines(x, y, lwd = 2)
text(-.5, .15, "p = .75", cex = 1.2)
text(qnorm(.75), .38, "?", cex = 1.2)
```

```{r}
qnorm(.75, mean = 0, sd = 1)
```

## Sampling from a Continuous Random Variable

For example $Y \sim \mbox{Normal}(0, 1)$

## Sampling from a Continuous Random Variable

For example $Y \sim \mbox{Normal}(0, 1)$

```{r}
rnorm(5, mean = 0, sd = 1)
```

## Trial & Error

- Plot the density function of a $\mbox{Normal}(-10, 10)$. What is a reasonable "zoom" for the plot and how do you achieve it?
- Draw a histogram (`hist()`) with 10000 realizations from a $\mbox{Normal}(-10, 10)$. How does it compare to the density function plot?
- Find the median of a standard normal distribution $\mbox{Normal}(0, 1)$ using `dnorm()`, `pnorm()`, or `qnorm()`

# Signal Detection Theory Model

## SDT model for subliminal perception

General idea: Perception strength varies gradually, but is on average higher when the stimulus is present compared to absent.

```{r echo = F}
x <- seq(-3, 5, .01)
y.noise <- dnorm(x)
y.signal <- dnorm(x, 1.5)

plot(x, y.noise
     , type = "l", lwd = 2
     , xlim = range(x)
     , frame.plot = F
     , ylab = "Density"
     , xlab = "Sensory Strength"
     )
lines(x, y.signal, col = "firebrick4", lwd = 2)
# make.line(0)
# make.line(1.5, 1.5)
# abline(v = 1, lwd = 2, col = "darkgreen")
# axis(3, at = c(0, 1.5), labels = c("", ""))
# mtext("d'", 3, line = .5, at = .75, cex = 1.3)
# text(1.2, .03, "c", cex = 1.3)
text(-2, .25, "Stimulus absent")
text(3.5, .25, "Stimulus present")
```

## SDT model for subliminal perception

General idea: Perception strength varies gradually, but is on average higher when the stimulus is present compared to absent.

```{r echo = F}
x <- seq(-3, 5, .01)
y.noise <- dnorm(x)
y.signal <- dnorm(x, 1.5)

plot(x, y.noise
     , type = "l", lwd = 2
     , xlim = range(x)
     , frame.plot = F
     , ylab = "Density"
     , xlab = "Sensory Strength"
     )
lines(x, y.signal, col = "firebrick4", lwd = 2)
# make.line(0)
# make.line(1.5, 1.5)
# abline(v = 1, lwd = 2, col = "darkgreen")
axis(3, at = c(0, 1.5), labels = c("", ""))
mtext("d'", 3, line = .5, at = .75, cex = 1.3)
# text(1.2, .03, "c", cex = 1.3)
text(-2, .25, "Stimulus absent")
text(3.5, .25, "Stimulus present")
```

## SDT model for subliminal perception

A criterion determines the response made.

```{r echo = F}
x <- seq(-3, 5, .01)
y.noise <- dnorm(x)
y.signal <- dnorm(x, 1.5)

plot(x, y.noise
     , type = "l", lwd = 2
     , xlim = range(x)
     , frame.plot = F
     , ylab = "Density"
     , xlab = "Sensory Strength"
     )
lines(x, y.signal, col = "firebrick4", lwd = 2)
# make.line(0)
# make.line(1.5, 1.5)
abline(v = 1, lwd = 2, col = "darkgreen")
axis(3, at = c(0, 1.5), labels = c("", ""))
mtext("d'", 3, line = .5, at = .75, cex = 1.3)
text(1.2, .03, "c", cex = 1.3)
text(-2, .25, "Stimulus absent")
text(3.5, .25, "Stimulus present")
```

## SDT model compared to HT model

One parameter for perception ability, one for response bias

>- $d'$ corresponds to $d$
>- $c$ corresponds to $g$

## SDT model predictions for H, M, F, C

What corresponds to the probability of hit? Area under the curve!

```{r echo = F}
x <- seq(-3, 5, .01)
y.noise <- dnorm(x)
y.signal <- dnorm(x, 1.5)

plot(x, y.noise
     , type = "l", lwd = 2
     , xlim = range(x)
     , frame.plot = F
     , ylab = "Density"
     , xlab = "Sensory Strength"
     )
lines(x, y.signal, col = "firebrick4", lwd = 2)
# make.line(0)
# make.line(1.5, 1.5)
abline(v = 1, lwd = 2, col = "darkgreen")
# axis(3, at = c(0, 1.5), labels = c("", ""))
# mtext("d'", 3, line = .5, at = .75, cex = 1.3)
text(1.2, .03, "c", cex = 1.3)
text(-2, .25, "Stimulus absent")
text(3.5, .25, "Stimulus present")
```

## SDT model predictions for H, M, F, C

What corresponds to the probability of hit? Area under the curve!

```{r echo = F}
x <- seq(-3, 5, .01)
y.noise <- dnorm(x)
y.signal <- dnorm(x, 1.5)

plot(x, y.noise
     , type = "l", lwd = 2
     , xlim = range(x)
     , frame.plot = F
     , ylab = "Density"
     , xlab = "Sensory Strength"
     )
lines(x, y.signal, col = "firebrick4", lwd = 2)
# make.line(0)
# make.line(1.5, 1.5)
abline(v = 1, lwd = 2, col = "darkgreen")
# axis(3, at = c(0, 1.5), labels = c("", ""))
# mtext("d'", 3, line = .5, at = .75, cex = 1.3)
text(1.2, .03, "c", cex = 1.3)
text(-2, .25, "Stimulus absent")
text(3.5, .25, "Stimulus present")
make.poly(1, 5, mu = 1.5, col = adjustcolor("firebrick", .3))
```

## SDT model predictions for H, M, F, C

What corresponds to the probability of miss?

```{r echo = F}
x <- seq(-3, 5, .01)
y.noise <- dnorm(x)
y.signal <- dnorm(x, 1.5)

plot(x, y.noise
     , type = "l", lwd = 2
     , xlim = range(x)
     , frame.plot = F
     , ylab = "Density"
     , xlab = "Sensory Strength"
     )
lines(x, y.signal, col = "firebrick4", lwd = 2)
# make.line(0)
# make.line(1.5, 1.5)
abline(v = 1, lwd = 2, col = "darkgreen")
# axis(3, at = c(0, 1.5), labels = c("", ""))
# mtext("d'", 3, line = .5, at = .75, cex = 1.3)
text(1.2, .03, "c", cex = 1.3)
text(-2, .25, "Stimulus absent")
text(3.5, .25, "Stimulus present")
make.poly(-3, 1, mu = 1.5, col = adjustcolor("firebrick", .3))
```

## SDT model predictions for H, M, F, C

What corresponds to the probability of false alarm?

```{r echo = F}
x <- seq(-3, 5, .01)
y.noise <- dnorm(x)
y.signal <- dnorm(x, 1.5)

plot(x, y.noise
     , type = "l", lwd = 2
     , xlim = range(x)
     , frame.plot = F
     , ylab = "Density"
     , xlab = "Sensory Strength"
     )
lines(x, y.signal, col = "firebrick4", lwd = 2)
# make.line(0)
# make.line(1.5, 1.5)
abline(v = 1, lwd = 2, col = "darkgreen")
# axis(3, at = c(0, 1.5), labels = c("", ""))
# mtext("d'", 3, line = .5, at = .75, cex = 1.3)
text(1.2, .03, "c", cex = 1.3)
text(-2, .25, "Stimulus absent")
text(3.5, .25, "Stimulus present")
make.poly(1, 5, mu = 0, col = adjustcolor(1, .3))
```

## SDT model predictions for H, M, F, C

What corresponds to the probability of correct rejection?

```{r echo = F}
x <- seq(-3, 5, .01)
y.noise <- dnorm(x)
y.signal <- dnorm(x, 1.5)

plot(x, y.noise
     , type = "l", lwd = 2
     , xlim = range(x)
     , frame.plot = F
     , ylab = "Density"
     , xlab = "Sensory Strength"
     )
lines(x, y.signal, col = "firebrick4", lwd = 2)
# make.line(0)
# make.line(1.5, 1.5)
abline(v = 1, lwd = 2, col = "darkgreen")
# axis(3, at = c(0, 1.5), labels = c("", ""))
# mtext("d'", 3, line = .5, at = .75, cex = 1.3)
text(1.2, .03, "c", cex = 1.3)
text(-2, .25, "Stimulus absent")
text(3.5, .25, "Stimulus present")
make.poly(-3, 1, mu = 0, col = adjustcolor(1, .3))
```

## SDT model predictions for H, M, F, C {.build}

Let the CDF of a standard normal be $\Phi(x)$. Then

\[Pr(\mbox{correct rejection}) = \Phi(c)\]

\[Pr(\mbox{false alarm}) = 1 - \Phi(c)\]

Probability of a miss is the probability of a normal with mean $d'$ and variance $1$ to be less than $c$. Can be written as $\Phi(c - d')$.

\[Pr(\mbox{miss}) = \Phi(c - d')\]

\[Pr(\mbox{hit}) = 1 - \Phi(c - d') \]

## SDT model in model form

\[
Y_h \sim \mbox{Binomial}( 1 - \Phi(c - d'), N_s),\\
Y_f \sim \mbox{Binomial}( 1 - \Phi(c), N_n).
\]

## SDT in R

```{r}
#log likelihood for signal detection
ll.sd <- function(par, y){       #par = c(d', c) y = c(hit, miss, fa, cr)
p <- 1:4
p[1] <- 1 - pnorm(par[2], par[1], 1)
p[2] <- 1 - p[1]
p[3] <- 1 - pnorm(par[2], 0, 1)
p[4] <- 1 - p[3]
-sum(y * log(p))
}
```

## SDT in R

```{r}
#log likelihood for signal detection
ll.sd <- function(par, y){       #par = c(d', c) y = c(hit, miss, fa, cr)
p <- 1:4
p[1] <- 1 - pnorm(par[2], par[1], 1)
p[2] <- 1 - p[1]
p[3] <- 1 - pnorm(par[2], 0, 1)
p[4] <- 1 - p[3]
-sum(y * log(p))
}
```

```{r}
y <- c(40, 10, 30, 20)
par <- c(1, 0) #starting values
out <- optim(par, ll.sd, y = y)
out$par
```

## SDT for 2 Conditions

12ms and 20ms presentation time

```{r echo = F}
x <- seq(-3, 6, .01)
y.noise <- dnorm(x)
y.signal <- dnorm(x, 1)
y.signal2 <- dnorm(x, 2.5)

plot(x, y.noise
     , type = "l", lwd = 2
     , xlim = range(x)
     , frame.plot = F
     , ylab = "Density"
     , xlab = "Sensory Strength"
     )
lines(x, y.signal, col = "firebrick4", lwd = 2)
lines(x, y.signal2, col = "darkblue", lwd = 2)
# make.line(0)
# make.line(1.5, 1.5)
abline(v = .8, lwd = 2, col = "darkgreen")
axis(3, at = c(0, 1, 2.5), labels = c("", "", ""))
mtext(expression("d'"[1]), 3, line = .5, at = .5, cex = 1.3)
mtext(expression("d'"[2]), 3, line = .5, at = 1.25, cex = 1.3)
text(.9, .03, "c", cex = 1.3)
text(-2, .25, "Stimulus absent")
text(.2, .15, "Cond 1")
text(3, .2, "Cond 2")
```

## SDT for 2 Conditions

```{r}
#par = c(d'1, d'2, c) y = c(hit1, miss1, fa1, cr1, hit2, miss2, fa2, cr2)
nll.sdt <- function(par3, y8){            
  ll.sd(par3[c(1, 3)], y8[1:4]) +     #condition 1
    ll.sd(par3[2:3], y8[5:8])       #condition 2
}

### Data analysis
dat <- c(22, 28, 22, 28   #h, m, f, c for condition 1
         , 35, 15, 21, 29) #h, m, f, c for condition 2

par.m <- c(1, 2, 1) #starting values
out2  <- optim(par.m, nll.sdt, y8 = dat, hessian = T)
```

## Do we need two separate d'?

Estimate a model with only one d' for both conditions.

## Do we need two separate d'?

Estimate a model with only one d' for both conditions.

```{r}
#par = c(d', c) y = c(hit1, miss1, fa1, cr1, hit2, miss2, fa2, cr2)
nll.sdt.2 <- function(par2, y8){            
  ll.sd(par2, y8[1:4]) +     #condition 1
    ll.sd(par2, y8[5:8])       #condition 2
}

### Data analysis
par.m <- c(1, .5) #starting values
out3  <- optim(par.m, nll.sdt.2, y8 = dat, hessian = T)
```

## Do we need two separate d'? | Model Comparison

```{r}
#G2 between general and constrained model
(G2 <- 2 * (out3$value - out2$value))

#p-value
1 - pchisq(G2, df = 1)
```




